Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-05-14 ebayes_rnaseq_heterosis.tex
thru 2.2

\renewcommand{\gamma}{c}

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.

To obtain estimates of the hyperparameters, we use \edgeR to provide estimates of gene-specific parameters, $\hat{\theta}$ and normalization factors, $\hat{\gamma}$.

Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 
\begin{equation}
\begin{split}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
&\times 
\left. \vphantom{\prod_{v=1}^V}
N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
\right]
\label{e:condind}
\end{split}
\end{equation}
Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.

As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate of $\theta$ and $\pi$. We explicitly use the MAP estimate of $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}. The MAP estimate of $\theta$ could be used as initial values in the MCMC, although we did not find this necessary.

@@

For a specific gene $g$, low-parent heteorisis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(LPH_g|y,\hat{\pi}_{MAP}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi}_{MAP}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous. With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in \RStan{}. Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via \edgeR{} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}. The {\tt Stan} software also includes the \RStan{} function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior. By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.

>>>>

%\renewcommand{\gamma}{c}

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.

To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values method of \cite{robinson2010scaling}. We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\phi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e. $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$. Using the gene-specific parameter estimates, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g. $\hat{\eta}_\delta = \sum_{g=1}^G \hat{\delta}_g/G$ and $2\hat{\sigma}_\delta^2 = \sum_{g=1}^G (\hat{\delta}_g - \hat{\eta}_\delta)^2/(G-1)$.

Conditional on the estimated normalization factors $\hat{\gamma}$ and hyperparameters $\hat{\pi}$, we perform a Bayesian analysis to re-estimate the gene-specific parameters and describe their uncertainty. Conditional on $\hat{\gamma}$ and $\hat{\pi}$ the gene-specific parameters are independent, i.e.
\begin{equation}
\begin{split}
p\left(\theta\left|y,\hat{\pi},\hat{\gamma}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
&\times 
\left. \vphantom{\prod_{v=1}^V}
N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
\right]
\label{e:condind}
\end{split}
\end{equation}
Thus, conditional posterior inference across the genes can be obtained independently and in parallel.

To perform this Bayesian analysis, we used the statistical software \Stan which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtaining samples from the posterior in equation \eqref{e:condind}. We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference. We monitored convergence using the potential scale reduction factor (PSRF) \citep{Gelm:Rubi:infe:1992} for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$. If the maximum PSRF was greater than 1.1, we reran the chain with an additional 500 iterations (250 burnin and 250 for inference) and repeated this process until all PSRF's were less than 1.1.

@@

For a specific gene $g$, low-parent heterosis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(LPH_g|y,\hat{\pi},\hat{\gamma}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

%To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous. With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in \RStan{}. Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via \edgeR{} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.
